Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-02-25 main.Rmd
nonlinear to cyclic

abstract: |
 Recent advances in technology greatly facilitates recording and storing data at much finer temporal scales than was previously possible. As the frequency of time-oriented data increases, the number of questions about the observed variable that can be addressed by visual representation also increases. We propose some new tools to explore this type of data, which deconstruct time in many different ways. There are several classes of time deconstructions including linear and nonlinear time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. Nonlinear time granularities can be circular such as hour of the day, and day of the week, quasi-circular such as day of the month, and aperiodic such as public or school holidays.

@@

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within a tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk].
bibliography: [bibliography.bib]
preamble: >
  \setlength {\marginparwidth }{2cm}
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

@@

Linear time granularities respect the linear progression of time such as hours, days, weeks and months, and non-repeating. One of the first attempts to characterize these granularities is due to @Bettini1998-ed. However, the definitions and rules defined are inadequate for describing nonlinear or repeating granularities. Hence, there is a need to define some new nonlinear time granularities, that can be useful in visualizations. Nonlinear time granularities can be circular, quasi-circular or aperiodic. Examples of circular granularities are hour of the day and day of the week; an examples of a quasi-circular granularity is day of the month; examples of aperiodic granularities are public holidays and school holidays.

@@

```{r linear-time, echo=FALSE, out.width = "100%", fig.cap="The time domain distributed as linear granularities."}
knitr::include_graphics("Figs/linear-arr.png")
```
In addition to the integer index, each granule could be associated with a textual representation which is called the label. The label can be mapped to index for input and the index can later be mapped to this label for output. When the label set is exactly the integers, the granularities are called "full-integer" labeled granularity, such as in Definition \ref{def:linear}.<!-- Some calendar operations will only work on full-integer labeled granularities, while others can be implemented using more flexible labelings.-->

@Bettini1998-ed discusses the relationships between linear time granularities, periodicities in time, and the structure of a calendar.

<!-- CONTENT CUT -->
<!-- \begin{definition}[Finer than]\label{def:finerthan} -->
<!-- A linear granularity $G$ is finer than a linear granularity $H$, denoted $G \preceq H$, if for each index $i$, there exists an index $j$ such that $G(i) \subset H(j)$. -->
<!-- \end{definition}. -->

@@

<!-- CONTENT CUT -->
<!-- These definitions are illustrated in \autoref{fig:groups-finer}, where both day $\trianglelefteq$ week, and day $\preceq$ week, hold true. However, weekends are finer than weeks but do not group into weeks. And weekends neither groups into or is finer than public holidays. Consider another example where day $\trianglelefteq$ month. This relationship is not fully described until we associate it with periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping (day, month) is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions. -->

For example, day $\trianglelefteq$ week since every granule of "week" is the union of some set of granules of "day". Consider another example where day $\trianglelefteq$ month. This relationship however is incomplete without its association to periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping (day, month) is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions.

<!-- CONTENT CUT -->
<!-- # ```{r groups-finer, echo=FALSE, out.width = "100%", fig.cap="Illustrations of groups into and finer than relationships. Weekends are finer than weeks but do not group into weeks. And weekends neither groups into or finer than public holidays"} -->
<!-- knitr::include_graphics("Figs/groupsandfiner.png") -->
<!-- ``` -->

@@

<!-- CONTENT CUT -->
<!-- A granularity $H$ which is periodical with respect to $G$ is specified by: -->
<!-- (i) the $R$ sets of indexes of $G$, ${S_0,\dots,S_{R-1}}$ describing the granules of $H$ within one period; -->
<!-- (ii) the value of $P$; -->
<!-- (iii) the indexes of first and last granules in $H$, if their value is not infinite. -->

If ${S_0,\dots,S_{R-1}}$ are the sets of indexes of $G$ describing
$H(0), \dots, H(R - 1)$, respectively, then the description of an arbitrary granule $H(j)$ is given by $\bigcup_{i \in S_j \mod R}G(P*\lfloor j/R \rfloor + i)$. Also, granularities can be periodical with respect to other granularities, except for a finite number of spans of time where they behave in an anomalous way [@Bettini2000-vy]. These are called quasi-periodic relationships. Intuitively, the definition requires that all granules of $H$ within the span of time between two exceptions have the same periodical behavior, characterized by $R$ and $P$.

@@

<!-- CONTENT CUT -->
<!-- \begin{definition}[Groups quasi-periodically]\label{def:quasi} -->
<!-- A granularity $H$ is quasi-periodical with respect to a granularity $G$ if: -->
<!-- (1) $G \trianglelefteq H$; and -->
<!-- (2) there exist a set of intervals $E_1,\dots,E_z$ (the granularity exceptions) and positive integers $R$, $P$, where $R$ is less than the minimum of the number of granules of $H$ between any two exceptions, such that for all $i \in \mathbb{Z}$, if $H(i) = \bigcup_{k \in [0,k]} G(j_r)$  and $H (i + R) \neq \phi$ and $i + R < min(E)$, where $E$ is the closest existing exception after $H (i)^2$, then $H (i + R) = \bigcup_{k \in [0,k]} G(j_r + P)$. -->
<!-- \end{definition} -->

@@

For example, in a Gregorian calendar without leap years each month is a grouping of the same number of days each year. The period is $P = 365$ and the number of granules of "months" in each period is given by $R=12$. Thus, we could say days group periodically into months. In a Gregorian calendar with leap years, days group quasi-periodically into months. The value of $P$ and $R$ would still be 365 and 12 respectively, while the <!--$E_i$s--> exceptions would be the time domain corresponding to $29^{\text{th}}$ February of any year.

<!-- CONTENT CUT -->
<!-- \begin{definition}[Bottom granularity]\label{def:B} -->
<!-- Given a granularity order relationship g-rel and a set of granularities having the same time domain, a granularity B in the set is a bottom granularity with respect to g-rel, if B g-rel H for each granularity H in the set. -->
<!-- \end{definition} -->

<!-- Here, g-rel could be any relationship defined between granularities as defined in @Bettini1998-ed. Given the set of all granularities defined over the time domain $(\mathbb{Z}; <)$, and the granularity relationship $\trianglelefteq$ (groups into), the granularity mapping each index into the corresponding instant (same integer number as the index) is a bottom granularity with respect to $\trianglelefteq$. -->

A discrete time model often uses a fixed smallest granularity also known as _bottom granularity_ (@Bettini1998-ed) so that the the underlying time domain can be described as a sequence of non-decomposable, consecutive time intervals of some fixed, minimal duration called _chronons_ (@Jensen1998-qn). Chronons may be grouped into granules and a non-overlapping mapping of granules to subsets of the time domain is a linear granularity. A system of multiple granularities in lattice structures is referred to as a _calendar_ (see @Dyreson2000-mr).

@@

Linear time granularities are computed through an algebraic representation for time granularities, which is referred to as calendar algebra [@Ning2002-tf]. It is assumed that there exists a *_bottom granularity_* and calendar algebra operations are designed to generate new granularities recursively from the bottom up. Some relevant calendar algebra operations are discussed below; these will be used in Section \ref{sec:cyclic-gran} for illustrations in circular and aperiodic granularities.

<!-- CONTENT CUT -->
<!-- The calendar algebra consists of two kinds of operations: grouping-oriented and granule-oriented operations. The grouping-oriented operations combine certain granules of a granularity together to form the granules of the new granularity. For example, consider a calendar with only two linear granularities: "minute" and "hour", where "hour" is generated by grouping every $60$ "minutes". -->
<!-- The granule-oriented operations do not change the granules of a granularity, but rather select which granules should remain in the new granularity. For example, one can define the granularity "Monday" and hence select only Mondays from linear granularity "day". -->

@@

<!-- CONTENT CUT -->
<!-- The notion of a labeled granularity is introduced in @Bettini2000-qk to simplify the calendar algebra. A labeled granularity is a pair $(L, G)$, where the label set $L$ is a subset of the integers instead of the set of integers and -->
<!-- follows the conditions (1) and (2) in \ref{def:linear} where $i$, $j$, and $k$ are in $L$. When the label set $L$ is exactly the integers, the granularities are called "full-integer" labeled granularity, such as in Definition \ref{def:linear}.<!-- However, the set $L$ of labels can be an arbitrary subset of (possibly non contiguous) integers and could be used to identify granules.-->

<!-- $G$ is a mapping from $L$ to the subsets of the time domain such that: -->

<!--   1. if $i<j$ , where $i$ and $j$ are in $L$, and $G(i)$ and $G(j)$ are non-empty, then each element in $G(i)$ is less than all the elements of $G(j)$; and -->
<!--   2. if $i<k<j$ , where $i$, $j$, and $k$ are in $L$, and $G(i)$ and $G(j)$ are non-empty, then $G(k)$ is non-empty. -->

The grouping operation:
: Let $G_1$ be a full-integer labeled granularity, and $m$ a positive integer. The grouping operation Group$_m(G)$ generates a new granularity $G_2$, by partitioning the granules of $G_1$ into $m$-granule groups and making each group a granule of the resulting granularity. More precisely, $G_2 = \text{Group}_m(G_1)$ is the full-integer labeled granularity such that for each integer $i$,
$$G_2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G_1(j).$$
Example: $\text{minute} = \text{Group}_{60}(\text{second})$; $\text{hour} = \text{Group}_{60}(\text{minute})$

<!-- CONTENT CUT -->
<!-- where $\text{second}(1)$ should coincide with $\text{minute}(1)$ and likewise. Otherwise, repeated operations of grouping might result in a granule of "resultant" granularity "day" or "week", which is different from a time-stamped "day" or "week". -->

@@

For example, due to the uneven length of months, we need to use the altering-tick operation to compute months from days.  Let us consider a granularity "pseudomonth" which computes months only for non-leap years by grouping 31 days, and then shrink April (4), June (6), September (9) and November (11) by 1 day, and shrink February (2) by 3 days [@Ning2002-tf].

\begin{align*}
\text{pseudomonth} = &\text{Alter}^{12}_{11,-1}(\text{day}, \text{Alter}^{12}_{9,-1}(\text{day}, \\
  &~~\text{Alter}^{12}_{6,-1}(\text{day}, \text{Alter}^{12}_{4,-1)}(\text{day}, \\
  &~~~\text{Alter}^{12}_{(2,-3)}(\text{day}, \text{Group}_{31}(\text{day})))))),
\end{align*}

For more variations of calendar algrabra operations, see @Ning2002-tf.

# Nonlinear time granularities {#sec:cyclic-gran}

@@

We propose a formalism of nonlinear time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. In a tsibble, each observation (row) is uniquely identified by index and key.

@@

Nonlinear time granularities with periodicities can be constructed by relating two linear granularities. The mappings between these linear granularities can be regular (minutes and hours) or irregular (days and months). <!--For example, a regular mapping exists between  (where 60 minutes always add up to 1 hour), while an irregular mapping exists between  (as the number of days per month ranges from 28 to 31).--> A grouping operation in Calendar algebra could lead to a regular mapping, whereas a altering-tick operation could result in an irregular mapping. Nonlinear time granularities are referred to as **circular** if mappings are regular and **quasi-circular** if the mapping is irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month.

@@

\autoref{fig:circular-dow} depicts the circular granularity "Day-of-Week", constructed from linear granularities "day" (the bottom granularity) and "week". The granules that overlap share elements from the underlying time domain. In this example, $P=7$, and the circular granularity $C_{\text{day}, \text{week}}$ representing day-of-week consists of the repeated pattern $\{L(0), L(1), L(2),$ $L(3), L(4), L(5), L(6)\}$.

@@

In general, any circular granularity relating two linear granularity can be expressed as $C_{(G, H)}(z) = L(\lfloor z/P(B,G) \rfloor\mod P(G,H))$, where linear granularity $H$ is periodic with respect to linear granularity $G$ with regular mapping such that the number of granules of $G$ in each granule of $H$ is $P(G,H)$. Table \ref{tab:definitions} shows representation of circular granularities $C_i$ relating two linear granularities with $P_i$ being the number of granules of the finer granularity in each granule of the coarser granularity and $L_i$ is the associated label mapping. <!--The representation of the resultant circular granularity will be a function of the index set.--> The bottom granularity is assumed to be minutes in this case.

@@

\begin{equation} \label{eq:eq4}
\begin{split}
Q_{B,G'}(8) & = L(8 - \sum_{w=0}^{3-1}\vert T_{w \mod 2}\vert) ,\quad since \quad 8 \in T_{3} \\
  & = L(1) \\
\end{split}
\end{equation}

<!-- CONTENT CUT -->
  <!-- & = L(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\ -->
  <!-- & = L(8 - \vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert - \vert T_{2 \mod 2}\vert) \\ -->
  <!-- & = L(8 - 2*\vert T_{0}\vert - \vert T_{1}\vert) \\ -->
  <!-- & = L(8 - 2*2 - 3) \\ -->

  <!-- & = L(10 - \sum_{w=0}^{0}\vert T_{w\mod 2}\vert) \\ -->
  <!-- & = L(10 - \vert T_{0 \mod 2}\vert) \\ -->
  <!-- & = L(10 - \vert T_{0}\vert) \\ -->
  <!-- & = L(10 - 7) \\ -->

\begin{equation} \label{eq:eq5}
\begin{split}
Q_{B,H'}(10) & = L(10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert) ,\quad since \quad  10 \in T_{1}  \\
  & = L(3) \\
\end{split}
\end{equation}

**Discussion**: If linear granularity $G'$ is periodical with respect to $B$ with irregular mapping, then there exist $R'$, $P' \in \mathbb{Z}_+$ such that if $G'(w) = \bigcup_{z \in T_w}B(z)$ then $$G'(w) = \bigcup_{z \in T_w \mod R'}B(P'*\lfloor w/R' \rfloor + z)$$ (from definition \ref{def:periodical})<!-- - $B \trianglelefteq G'$ --> <!-- and,  number of granules of $B$ in each granule of $G'$ is not a constant (irregular mapping) -->. Here $w \mod R'$ represents the index that must be shifted to obtain $G'(w)$. The idea here is if we know the composition of each of the granules of $G'$ in terms of granules of $B$ for one period, we can find the composition of any granule of $G'$ beyond a period since the "pattern" repeats itself along the time domain due to the periodic property. The periodic property also ensures that $\vert T_w \vert = \vert T_{w\mod R'} \vert$ since every $w^{\text{th}}$ and $(w+R')^{\text{th}}$ granule of $G'$ will have the same number of granules of $B$. The term $\sum_{w=0}^{k-1}\vert T_{w}\vert$ denotes the number of granules of $B$ till the $(k-1)^{\text{th}}$ granule of $G'$. Since $\vert T_w \vert = \vert T_{w\mod R'} \vert$, the number of granules of $B$ till the $(k-1)^{\text{th}}$ granule of $G'$ becomes $\sum_{w=0}^{k-1}\vert T_{w \mod R'}\vert$ in Definition \autoref{def:quasicircular}.

@@

It can be noted here that if a linear granularity $G'$ is quasi-periodic (Bettini2000-vy) <!--(Definition \autoref{def:quasi})--> with respect to $B$, then \autoref{eq:quasi} can be modified as follows to account for exceptions $E = [e_{begin}, e_{end}]$.<!--(by Definition \autoref{def:quasi}).-->

@@

Periodic and Quasi-periodic behavior can be defined by a repeating pattern, except for a finite number of granules that can be seen as discontinuity points in the granularity in case of quasi-periodic behavior. Aperiodic time granularities are the ones which can not be specified as a periodic repetition of a pattern of granules. Most public holidays repeat every year, but there is no finite (or reasonably small) period within which their behavior remains constant. A classic example can be that of Easter, whose dates repeat only after 5,700,000 years. U.S labour day is the first Monday in September and U.S. Memorial Day is the last day in May. In Australia, if a standard public holiday falls on a weekend, a substitute public holiday will sometimes be observed on the first non-weekend day (usually Monday) after the weekend. Examples of aperiodic granularity may also include school holidays or a scheduling event that might cover the first and third Monday of the month between June and October, except for state holidays. All of these are recurring events, but with non-periodic patterns. As such, plausible $P_i$ from \autoref{tab:quasi} could be possibly infinite for aperiodic granularities.

Aperiodic nonlinear granularities are defined using aperiodic linear granularities. Consider n aperiodic linear granularities $M_i$  $\forall \{i\in{1, 2, \dots, n\}}$, that is, none of them can be expressed as periodic or quasi-periodic with respect to the bottom granularity. Further assume that $B \trianglelefteq M_i$ $\forall \{i\in{1, 2, \dots, n\}}$. Then according to Definition \ref{def:groupsinto}, for each index $j$ there exists a (possibly infinite) subset $T_{\{i_j\}}$ of the integers such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$ and suppose $M = \bigcup_{i=1}^{n}M_{i}$ is formed by collecting the granules of $\{M_1, M_2, \dots, M_n\}$. Here, index ${\{i_j\}}$ stands for the $j^{\text{th}}$ granule of the $i^{\text{th}}$ linear aperiodic granularity.

>>>>

abstract: |
 Recent advances in technology greatly facilitates recording and storing data at much finer temporal scales than was previously possible. As the frequency of time-oriented data increases, the number of questions about the observed variable that can be addressed by visual representation also increases. We propose some new tools to explore this type of data, which deconstruct time in many different ways. There are several classes of time deconstructions including linear and cyclic time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. Cyclic time granularities can be circular such as hour of the day, and day of the week, quasi-circular such as day of the month, and aperiodic such as public or school holidays.

@@

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within a tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk].
bibliography: [bibliography.bib]
preamble: >
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

@@

Linear time granularities respect the linear progression of time such as hours, days, weeks and months, and non-repeating. One of the first attempts to characterize these granularities is due to @Bettini1998-ed. However, the definitions and rules defined are inadequate for describing cyclic or repeating granularities. Hence, there is a need to define some new cyclic time granularities, that can be useful in visualizations. Cyclic time granularities can be circular, quasi-circular or aperiodic. Examples of circular granularities are hour of the day and day of the week; an examples of a quasi-circular granularity is day of the month; examples of aperiodic granularities are public holidays and school holidays.

@@

```{r linear-time, echo=FALSE, out.width = "100%", fig.cap="The time domain distributed as linear granularities."}
knitr::include_graphics("Figs/linear-arr.png")
```

@Bettini1998-ed discusses the relationships between linear time granularities, the structure of a calendar, and periodicities in time.

\begin{definition}[Finer than]\label{def:finerthan}
A linear granularity $G$ is finer than a linear granularity $H$, denoted $G \preceq H$, if for each index $i$, there exists an index $j$ such that $G(i) \subset H(j)$.
\end{definition}.

@@

These definitions are illustrated in \autoref{fig:groups-finer}, where both day $\trianglelefteq$ week, and day $\preceq$ week, hold true. However, weekends are finer than weeks but do not group into weeks. And weekends neither groups into or is finer than public holidays. Consider another example where day $\trianglelefteq$ month. This relationship is not fully described until we associate it with periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping (day, month) is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions.

```{r groups-finer, echo=FALSE, out.width = "100%", fig.cap="Illustrations of groups into and finer than relationships. Weekends are finer than weeks but do not group into weeks. And weekends neither groups into or finer than public holidays"}
knitr::include_graphics("Figs/groupsandfiner.png")
```

@@

A granularity $H$ which is periodical with respect to $G$ is specified by:
(i) the $R$ sets of indexes of $G$, ${S_0,\dots,S_{R-1}}$ describing the granules of $H$ within one period;
(ii) the value of $P$;
(iii) the indexes of first and last granules in $H$, if their value is not infinite.
Then, if ${S_0,\dots,S_{R-1}}$ are the sets of indexes of $G$ describing
$H(0), \dots, H(R - 1)$, respectively, then the description of an arbitrary granule $H(j)$ is given by $\bigcup_{i \in S_j \mod R}G(P*\lfloor j/R \rfloor + i)$. Also, granularities can be periodical with respect to other granularities, except for a finite number of spans of time where they behave in an anomalous way [@Bettini2000-vy].

@@

\begin{definition}[Groups quasi-periodically]\label{def:quasi}
A granularity $H$ is quasi-periodical with respect to a granularity $G$ if:
(1) $G \trianglelefteq H$; and
(2) there exist a set of intervals $E_1,\dots,E_z$ (the granularity exceptions) and positive integers $R$, $P$, where $R$ is less than the minimum of the number of granules of $H$ between any two exceptions, such that for all $i \in \mathbb{Z}$, if $H(i) = \bigcup_{k \in [0,k]} G(j_r)$  and $H (i + R) \neq \phi$ and $i + R < min(E)$, where $E$ is the closest existing exception after $H (i)^2$, then $H (i + R) = \bigcup_{k \in [0,k]} G(j_r + P)$.
\end{definition}

Intuitively, the definition requires that all granules of $H$ within the span of time between two exceptions have the same periodical behavior, characterized by $R$ and $P$.

@@

Consider the Gregorian calendar without leap years. Then each month is a grouping of the same number of days each year. The period is $P = 365$ and the number of granules of "months" in each period is given by $R=12$. Thus, we could say days groups periodically into months. Now consider the Gregorian calendar with leap years, so that days group quasi-periodically into months. The value of $P$ and $R$ would still be 365 and 12 respectively, while the $E_i$s would be the time domain corresponding to $29^{\text{th}}$ February of any year.

\begin{definition}[Bottom granularity]\label{def:B}
Given a granularity order relationship g-rel and a set of granularities having the same time domain, a granularity B in the set is a bottom granularity with respect to g-rel, if B g-rel H for each granularity H in the set.
\end{definition}

Here, g-rel could be any relationship defined between granularities like "groups into", "finer than" or "groups periodically". Given the set of all granularities defined over the time domain $(\mathbb{Z}; <)$, and the granularity relationship $\trianglelefteq$ (groups into), the granularity mapping each index into the corresponding instant (same integer number as the index) is a bottom granularity with respect to $\trianglelefteq$.

<!-- Most practical problems seem to require only a granularity system containing a set of time granularities which are all periodical (or quasi-periodical) with respect to a basic granularity. The assumption is there should be one bottom granularity and all the other granularities are generated from the bottom granularity by calendar algebraic operations. -->

@@

Linear time granularities are computed through an algebraic representation for time granularities, which is referred to as calendar algebra [@Ning2002-tf]. It is assumed that there exists a “bottom” granularity and calendar algebra operations are designed to generate new granularities recursively from the bottom up. The calendar algebra consists of two kinds of operations: grouping-oriented and granule-oriented operations. The grouping-oriented operations combine certain granules of a granularity together to form the granules of the new granularity. For example, consider a calendar with only two linear granularities: "minute" and "hour", where "hour" is generated by grouping every $60$ "minutes". The granule-oriented operations do not change the granules of a granularity, but rather select which granules should remain in the new granularity. For example, one can define the granularity "Monday" and hence select only Mondays from linear granularity "day".

<!-- It might be the case that we do not always have integers as the label sets of granularities.  -->

@@

The notion of a labeled granularity is introduced in @Bettini2000-qk to simplify the calendar algebra. A labeled granularity is a pair $(L, G)$, where $L$ is a subset of the integers, and $G$ is a mapping from $L$ to the subsets of the time domain such that:

  1. if $i<j$ , where $i$ and $j$ are in $L$, and $G(i)$ and $G(j)$ are non-empty, then each element in $G(i)$ is less than all the elements of $G(j)$; and
  2. if $i<k<j$ , where $i$, $j$, and $k$ are in $L$, and $G(i)$ and $G(j)$ are non-empty, then $G(k)$ is non-empty.

$L$ is called the label set of $G$. When $L$ is exactly the integers, the granularities are called "full-integer" labeled granularity, such as in Definition \ref{def:linear}. However, the set $L$ of labels can be an arbitrary subset of (possibly non contiguous) integers and could be used to identify granules. Some calendar operations will only work on full-integer labeled granularities, while others can be implemented using more flexible labelings.

Some relevant grouping-oriented operations are discussed below; these will be used in Section \ref{sec:cyclic-gran} to define circular and aperiodic granularities.

The grouping operation:
: Let $G_1$ be a full-integer labeled granularity, and $m$ a positive integer. The grouping operation Group$_m(G)$ generates a new granularity $G_2$, by partitioning the granules of $G_1$ into $m$-granule groups and making each group a granule of the resulting granularity. More precisely, $G_2 = \text{Group}_m(G_1)$ is the full-integer labeled granularity such that for each integer $i$,
$$G_2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G_1(j).$$
For example:
\begin{align*}
  \text{minute} &= \text{Group}_{60}(\text{second});\\
  \text{hour}   &= \text{Group}_{60}(\text{minute});\\
  \text{day}    &= \text{Group}_{24}(\text{hour});\\
  \text{week}   &= \text{Group}_{7}(\text{day});
\end{align*}
where $\text{second}(1)$ should coincide with $\text{minute}(1)$ and likewise. Otherwise, repeated operations of grouping might result in a granule of "resultant" granularity "day" or "week", which is different from a time-stamped "day" or "week".

@@

For example, due to the uneven length of months, we need to use the altering-tick operation to compute months from days.  Let us consider a granularity "pseudomonth" which computes months only for non-leap years.
\begin{align*}
\text{pseudomonth} = &\text{Alter}^{12}_{11,-1}(\text{day}, \\
  &~~\text{Alter}^{12}_{9,-1}(\text{day}, \\
  &~~~~\text{Alter}^{12}_{6,-1}(\text{day}, \\
  &~~~~~~\text{Alter}^{12}_{4,-1)}(\text{day}, \\
  &~~~~~~~~\text{Alter}^{12}_{(2,-3)}(\text{day}, \\
  &~~~~~~~~~~\text{Group}_{31}(\text{day})))))),
\end{align*}
where the granularity pseudomonth is generated by grouping 31 days, and then shrink April (4), June (6), September (9) and November (11) by 1 day, and shrink February (2) by 3 days [@Ning2002-tf].

For more variations of grouping operations and granule oriented operations, see @Ning2002-tf.

# cyclic time granularities {#sec:cyclic-gran}

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. In a tsibble, each observation (row) should be uniquely identified by index and key.

@@

cyclic time granularities with periodicities can be constructed by relating two linear granularities. The mappings between these linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours (where 60 minutes always add up to 1 hour), while an irregular mapping exists between days and months (as the number of days per month ranges from 28 to 31). Mappings will be regular if linear granularities are formed by a grouping operation, but granularities formed by the altering-tick operation will be irregular. Cyclic time granularities are referred to as **circular** if mappings between linear time granularities are regular and **quasi-circular** if the mapping is irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month.

@@

\autoref{fig:circular-dow} depicts the circular granularity "day of the week", constructed from two linear granularities "day" (the bottom granularity) and "week". The granules that overlap share elements from the underlying time domain. In this example, $P=7$, and the circular granularity $C_{\text{day}, \text{week}}$ representing day-of-week consists of the repeated pattern $\{L(0), L(1), L(2),$ $L(3), L(4), L(5), L(6)\}$.

@@

In general, any circular granularity relating two linear granularity can be expressed as $C_{(G, H)}(z) = L(\lfloor z/P(B,G) \rfloor\mod P(G,H))$, where linear granularity $H$ is periodic with respect to linear granularity $G$ with regular mapping such that the number of granules of $G$ in each granule of $H$ is $P(G,H)$. Table \ref{tab:definitions} shows representation of circular granularities $C_i$ relating two linear granularities with $P_i$ being the number of granules of the finer granularity in each granule of the coarser granularity and $L_i$ is the associated label mapping. It is possible that none of the two linear granularities are bottom granularities. But the representation of the resultant circular granularity will be a function of the index set. The bottom granularity is assumed to be minutes.

@@

\begin{equation} \label{eq:eq4}
\begin{split}
Q_{B,G'}(8) & = L(8 - \sum_{w=0}^{3-1}\vert T_{w \mod 2}\vert) ,\quad since \quad 8 \in T_{3} \\
  & = L(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = L(8 - \vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert - \vert T_{2 \mod 2}\vert) \\
  & = L(8 - 2*\vert T_{0}\vert - \vert T_{1}\vert) \\
  & = L(8 - 2*2 - 3) \\
  & = L(1) \\
\end{split}
\end{equation}

\begin{equation} \label{eq:eq5}
\begin{split}
Q_{B,H'}(10) & = L(10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert) ,\quad since \quad  10 \in T_{1}  \\
  & = L(10 - \sum_{w=0}^{0}\vert T_{w\mod 2}\vert) \\
  & = L(10 - \vert T_{0 \mod 2}\vert) \\
  & = L(10 - \vert T_{0}\vert) \\
  & = L(10 - 7) \\
  & = L(3) \\
\end{split}
\end{equation}

**Discussion**: If linear granularity $G'$ is periodical with respect to $B$ with irregular mapping, then the following holds true:

- $B \trianglelefteq G'$
- there exist $R'$, $P' \in \mathbb{Z}_+$ such that if $G'(w) = \bigcup_{z \in T_w}B(z)$ then $$G'(w) = \bigcup_{z \in T_w \mod R'}B(P'*\lfloor w/R' \rfloor + z)$$

from definition \ref{def:periodical}

and,  number of granules of $B$ in each granule of $G'$ is not a constant (irregular mapping)

Here $w \mod R'$ represents the index that must be shifted to obtain $G'(w)$. The idea here is if we know the composition of each of the granules of $G'$ in terms of granules of $B$ for one period, we can find the composition of any granule of $G'$ beyond a period since the "pattern" repeats itself along the time domain due to the periodic property. The periodic property also ensures that $\vert T_w \vert = \vert T_{w\mod R'} \vert$ since every $w^{\text{th}}$ and $(w+R')^{\text{th}}$ granule of $G'$ will have the same number of granules of $B$. The term $\sum_{w=0}^{k-1}\vert T_{w}\vert$ denotes the number of granules of $B$ till the $(k-1)^{\text{th}}$ granule of $G'$. Since $\vert T_w \vert = \vert T_{w\mod R'} \vert$, the number of granules of $B$ till the $(k-1)^{\text{th}}$ granule of $G'$ becomes $\sum_{w=0}^{k-1}\vert T_{w \mod R'}\vert$ in Definition \autoref{def:quasicircular}.

@@

It can be noted here that if a linear granularity $G'$ is quasi-periodic (Definition \autoref{def:quasi}) with respect to $B$, then \autoref{eq:quasi} can be modified as follows to account for exceptions $E = [e_{begin}, e_{end}]$ (by Definition \autoref{def:quasi}).

@@

Periodic and Quasi-periodic behavior can be defined by a repeating pattern, except for a finite number of granules that can be seen as discontinuity points in the granularity in case of quasi-periodic behavior. Aperiodic time granularities are the ones which can not be specified as a periodic repetition of a pattern of granules. Most public holidays repeat every year, but there is no finite (or reasonably small) period within which their behavior remains constant. A classic example can be that of Easter, whose dates repeat only after 5,700,000 years. U.S labor day is the first Monday in September and U.S. Memorial Day is the last day in May. In Australia, if a standard public holiday falls on a weekend, a substitute public holiday will sometimes be observed on the first non-weekend day (usually Monday) after the weekend. Examples of aperiodic granularity may also include school holidays or a scheduling event that might cover the first and third Monday of the month between June and October, except for state holidays. All of these are recurring events, but with non-periodic patterns. As such, plausible $P_i$ from \autoref{tab:quasi} could be possibly infinite for aperiodic granularities.

Aperiodic cyclic granularities are defined using aperiodic linear granularities. Consider n aperiodic linear granularities $M_i$  $\forall \{i\in{1, 2, \dots, n\}}$, that is, none of them can be expressed as periodic or quasi-periodic with respect to the bottom granularity. Further assume that $B \trianglelefteq M_i$ $\forall \{i\in{1, 2, \dots, n\}}$. Then according to \autoref{def:groupsinto}, for each index $j$ there exists a (possibly infinite) subset $T_{\{i_j\}}$ of the integers such that
\begin{equation*}
M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)
\end{equation*}
and suppose $M = \bigcup_{i=1}^{n}M_{i}$ is formed by collecting the granules of $\{M_1, M_2, \dots, M_n\}$. Here, index ${\{i_j\}}$ stands for the $j^{\text{th}}$ granule of the $i^{\text{th}}$ linear aperiodic granularity.

\begin{definition}[Aperiodic cyclic granularity]\label{aperiodic}
An aperiodic cyclic granularity $A_{B, M}$ relates aperiodic linear granularity $M$ and bottom granularity $B$, if
\begin{equation}\label{eq:aperiodic}
\begin{split}
A_{B, M}(z) & = L(i), \quad z \in T_{i_j} \quad and \quad \forall i \in {1, 2, \dots, n}\\
     & = L(0), \quad otherwise
\end{split}
\end{equation}
where,
$z \in \mathbb{Z}_{\geq 0}$ denotes the index set,
$T_{i_j}$ are the sets of indices of $B$ describing aperiodic linear granularities $M_{i}(j)$ such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$, $M = \bigcup_{i=1}^{n}M_{i}$.
\end{definition}
